### 2024-05-19 thesis/input_files/data_exploratory.tex
May 19th 2024

\section{GEFCom2014}
The first class of datasets we identified were the ones from the GEFCom 2014 competition. The data is freely hosted on Dr.\ Hong blog \cite{hong2016probabilistic}.
The main reason is that, these datasets are considered an excellent test case for comparing predictive models between the EF community. Additionally, the scores of the competing models are freely available, this enables us to carry out a clean and transparent comparative study.
The GEFCom 2014 consisted of four tracks: price, load, wind power and solar power forecasting. 
The GEFCom2014 folder contains a zip file for each of the four tracks.
This thesis work is focused on providing forecasts for the load and price quantities.
\\
\subsection{GEFCom2014 Price track}
In the price forecasting track, the goal was to predict the electricity price for the next 24 hours of a single zone. 
The data can be found in the GEFCom2014-P\_V2 zip file alongside a set of instructions and the benchmark forecasts.
This data consisted of time series for the locational marginal price, for the zonal load and for the system load. The data covers the time interval ranging from the 1st of January 2011 to the 15th of June 2013, see figure \ref{fig:price_track_fig1}. 
Additionally, as the competition went on, the real observed data of the previous tasks were made available.
The fifteen target days ordered by task number are: 16/06/2013, 17/06/2013, 24/06/2013, 04/07/2013, 09/07/2013, 13/07/2013, 16/07/2013, 18/07/2013, 19/07/2013, 20/07/2013, 24/07/2013, 25/07/2013, 07/12/2013, 08/12/2013. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/price_track_fig1.png}
    \caption{Price track GEFCom2014}
    \label{fig:price_track_fig1}
\end{figure}
\subsubsection{EDA}
In exploring this data we started off by plotting the time series of the zonal prices alongside with the total load and zonal load, see figure \ref{fig:price_track_fig1}.
After that, we plotted the scatter plot for inspecting the correlation between the predictors and the dependent variables, figure \ref{fig:gefcom_zonal_price_vs_zonal_load} and figure \ref{fig:gefcom_zonal_price_vs_total_load}. From these visualisations, we can see their goodness for predicting zonal price since there exists a clear correlation between the candidate predictors and the target variables.

@@

\subsection{Load track}
In the load track, contestant were asked to provide one month ahead hourly probabilistic forecasts on a rolling basis for 15 consecutive rounds. To get started, in the first round, organizers provided 69 months of hourly load data (from 01/01/2005 to 30/10/2010) and 117 months of hourly temperature data (from 01/01/2001 to 30/10/2010). 
As for the price track, the true observed data from the previous track were made available as the competition progressed.
\\
The data for load forecasting is contained in the GEFCom2014-L\_V2 zip file. Within this subfolder, we have a txt file with the competition instructions and 15 subfolders one for each of the consecutive tasks. For each of those we have the prediction of the competition benchmark model and the train file upon which we will fit our model.
Any data from the various task folders are shifted by one month between eachother.
\\
In order to compare our model performance with the winning entries of the GEFCom2014 load competition, we will refer to the Provisional\_Leaderboard\_V2.csv file contained also in the GEFCom2014 Data directory.
% Notice a quick premise, within this xsls file the authors opted for various scores to compare the entries performance. This choice was motivated by the need 
% to give a proportial weight to each task when averaging between them; that is in order to not penalise too much the tasks with higher point spreads. Furthermore, additional factors such as frequency of benchmark outperformance, and robustness and clarity in methodology explaination.
\\
Notice, the pinball scores for the load track are stored inside the subtab L-score-0/L-score-2 of Provisional\_Leaderboard\_V2.
\subsubsection{EDA}
We applied the same kind of EDA to the GEFCom load data. Figure \ref{fig:gefcom_load_historical} depicts the hisorical load while figure \ref{fig:gefcom_w_avg_historical} reports the historical weather temperature. Finally, figure \ref{fig:gefcom_load_vs_w_avg} shows the scatter plot between the the load and the weather.
These plots motivate the selection of weather temperature as one of the predictors for the load.

>>>>

\section{GEFCom2014}
The first class of datasets we identified were the ones from the GEFCom 2014 competition. The data is freely hosted on Dr.\ Hong blog \cite{hong2016probabilistic}.
The main reason is that, these datasets are considered an excellent test case for comparing predictive models between the EF community. Additionally, the scores of the competing models are freely available, this enables us to carry out a clean and transparent comparative study.
The GEFCom 2014 consisted of four tracks: price, load, wind power and solar power forecasting. 
The GEFCom2014 folder contains a zip file for each of the four tracks.
This thesis work is focused on providing forecasts for the load and price quantities.
\\
\subsection{GEFCom2014 price track}
In the price forecasting track, the goal was to predict the electricity price for the next 24 hours of a single zone. 
The data can be found in the GEFCom2014-P\_V2 zip file alongside a set of instructions and the benchmark forecasts.
This data consisted of time series for the locational marginal price, for the zonal load and for the system load. The data covers the time interval ranging from the 1st of January 2011 to the 15th of June 2013, see figure \ref{fig:price_track_fig1}. 
Additionally, as the competition went on, the real observed data of the previous tasks were made available.
The fifteen target days ordered by task number are: 16/06/2013, 17/06/2013, 24/06/2013, 04/07/2013, 09/07/2013, 13/07/2013, 16/07/2013, 18/07/2013, 19/07/2013, 20/07/2013, 24/07/2013, 25/07/2013, 07/12/2013, 08/12/2013. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/price_track_fig1.png}
    \caption{Price track GEFCom2014}
    \label{fig:price_track_fig1}
\end{figure}
\subsubsection{EDA}
In exploring this data we started off by plotting the time series of the zonal prices alongside with the total load and zonal load, see figure \ref{fig:price_track_fig1}.
After that, we plotted the scatter plot for inspecting the correlation between the predictors and the dependent variables, figure \ref{fig:gefcom_zonal_price_vs_zonal_load} and figure \ref{fig:gefcom_zonal_price_vs_total_load}. From these visualisations, we can see their goodness for predicting zonal price since there exists a clear correlation between the candidate predictors and the target variables.

@@

\subsection{GEFCom2014 load track}
In the load track, contestant were asked to provide one month ahead hourly probabilistic forecasts on a rolling basis for 15 consecutive rounds. To get started, in the first round, organizers provided 69 months of hourly load data (from 01/01/2005 to 30/10/2010) and 117 months of hourly temperature data (from 01/01/2001 to 30/10/2010). 
As for the price track, the true observed data from the previous track were made available as the competition progressed.
\\
The data for load forecasting is contained in the GEFCom2014-L\_V2 zip file. Within this subfolder, we have a txt file with the competition instructions and 15 subfolders one for each of the consecutive tasks. For each of those we have the prediction of the competition benchmark model and the train file upon which we will fit our model.
Any data from the various task folders are shifted by one month between eachother.
\\
In order to compare our model performance with the winning entries of the GEFCom2014 load competition, we will refer to the Provisional\_Leaderboard\_V2.csv file contained also in the GEFCom2014 Data directory.
% Notice a quick premise, within this xsls file the authors opted for various scores to compare the entries performance. This choice was motivated by the need 
% to give a proportial weight to each task when averaging between them; that is in order to not penalise too much the tasks with higher point spreads. Furthermore, additional factors such as frequency of benchmark outperformance, and robustness and clarity in methodology explaination.
\\
Notice, the pinball scores for the load track are stored inside the subtab L-score-0/L-score-2 of Provisional\_Leaderboard\_V2.
\subsubsection{EDA}
We applied the same kind of EDA to the GEFCom load data. Figure \ref{fig:gefcom_load_historical} depicts the hisorical load while figure \ref{fig:gefcom_w_avg_historical} reports the historical weather temperature. Finally, figure \ref{fig:gefcom_load_vs_w_avg} shows the scatter plot between the the load and the weather.
These plots motivate the selection of weather temperature as one of the predictors for the load.

### 2024-05-19 thesis/input_files/implementation.tex
May 19th 2024

\section{Probabilistic forecasting}
\subsection{Linear quantile regression}
The implementation of \href{https://www.statsmodels.org/dev/generated/statsmodels.regression.quantile_regression.QuantReg.html}{quantile\_regression.QuantReg} from the regression module of statsmodels has been used.
The model is fitted through iterative reweighted least squares.
\subsection{Quantile gradient boosting machine}
The implementation of \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.GradientBoostingRegressor.html}{GradientBoostingRegressor} from the sklearn.ensemble submodule has been used.
\subsection{Quantile forest}
The implementation of \href{https://pypi.org/project/quantile-forest/}{quantile\_forest.RandomForestQuantileRegressor} has been used. This estimator is compatible with scikit-learn API \cite{Johnson2024}.
\subsection{Kernel quantile regression}
Kernel quantile regression had no previously implemented python open source library, thus the need of implementing our own version.
\\
The scikit-learn team provides a project template for the creation of estimators compatible with scikit-learn functionalities. Therefore, the KQR class is derived from the scikit-learn BaseEstimator and the mixin class RegressorMixin.
Our KQR class is initialized by providing a quantile, the regularization term C, the kernel family and its corresponding hyperparameters.
\\
In the fit method, we set up and solve the convex optimisation problem by using the cvxopt library, see offical manual \cite{vandenberghe2010cvxopt} for a reference.
When using this library, it is important to keep two things in mind. First this library assumes the quadratic term of the optimisation problem to be multiplied by the 0.5 factor, thus we just have to provide the Q matrix with no 0.5 in front.
Secondly, in order to specify multiple inequalities we have to stack them and provide them as a unique matrix.
\\
Once a solution to the convex problem has been found, we create a mask for the support vectors of the estimator in order to estimate the constant term of our kernel quantile regressor.
\\
In the predict method, we pass a matrix $X\_eval$ of independent variables, next we compute the kernel matrix between $X\_train$ and $X\_eval$ and obtain $y\_eval$ with the formula $y=\alpha^\intercal K+b$.
\\
This estimator is compatible with useful scikit-learn in built methods like gridsearch, crossvalidation and scoring rules. Moreover, this code is compatible with sklearn in built kernel functions \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.metrics.pairwise}{sklearn.metrics.pairwise} and \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.gaussian_process.kernels}{sklearn.gaussian\_process.kernels}, functions supported by our kernel quantile regression include: gaussian rbf, absolute laplacian, mater 1/2, matern 3/2, matern 5/2, linear, cosine, sigmoid, periodic, polynomial and custom composition of kernels.
% NOTICE, these kernel implementations differ in the form of passed parameters in the former we pass in the numerator gamma while in the latter we pass l, which stands for length scale in the denominator. To cross validate overt the same range of parameter across the different kernels we made sura that gamma=1/l.

>>>>

\section{Probabilistic forecasting}
\subsection{Linear quantile regression}
The implementation of \href{https://www.statsmodels.org/dev/generated/statsmodels.regression.quantile_regression.QuantReg.html}{quantile\_regression.QuantReg} from the regression module of statsmodels has been used.
The model is fitted through iterative reweighted least squares.
\subsection{Quantile gradient boosting machine}
The implementation of \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.GradientBoostingRegressor.html}{GradientBoostingRegressor} from the sklearn.ensemble submodule has been used.
\subsection{Quantile forest}
The implementation of \href{https://pypi.org/project/quantile-forest/}{quantile\_forest.RandomForestQuantileRegressor} has been used. This estimator is compatible with scikit-learn API \cite{Johnson2024}.
\subsection{Kernel quantile regression}
Kernel quantile regression had no previously implemented python open source library, thus the need of implementing our own version.
\\
The scikit-learn team provides a project template for the creation of estimators compatible with scikit-learn functionalities. Therefore, the KQR class is derived from the scikit-learn BaseEstimator and the mixin class RegressorMixin.
Our KQR class is initialized by providing a quantile, the regularization term C, the kernel family and its corresponding hyperparameters.
\\
In the fit method, we set up and solve the convex optimisation problem through the interior point method. This algorithm is taken from the cvxopt library, see its offical manual \cite{vandenberghe2010cvxopt} for a reference.
When using this library, it is important to keep two things in mind. First this library assumes the quadratic term of the optimisation problem to be multiplied by the 0.5 factor, thus we just have to provide the Q matrix with no 0.5 in front.
Secondly, in order to specify multiple inequalities we have to stack them and provide them as a unique matrix.
\\
Once a solution to the convex problem has been found, we create a mask for the support vectors of the estimator in order to estimate the constant term of our kernel quantile regressor.
\\
In the predict method, we pass a matrix $X\_eval$ of independent variables, next we compute the kernel matrix between $X\_train$ and $X\_eval$ and obtain $y\_eval$ with the formula $y=\alpha^\intercal K+b$.
\\
This estimator is compatible with useful scikit-learn in built methods like gridsearch, crossvalidation and scoring rules. Moreover, this code is compatible with sklearn in built kernel functions \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.metrics.pairwise}{sklearn.metrics.pairwise} and \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.gaussian_process.kernels}{sklearn.gaussian\_process.kernels}, functions supported by our kernel quantile regression include: gaussian rbf, absolute laplacian, mater 1/2, matern 3/2, matern 5/2, linear, cosine, sigmoid, periodic, polynomial and custom composition of kernels.
% NOTICE, these kernel implementations differ in the form of passed parameters in the former we pass in the numerator gamma while in the latter we pass l, which stands for length scale in the denominator. To cross validate overt the same range of parameter across the different kernels we made sura that gamma=1/l.
